---
title: "Fitting state-space models using the R package TMB"
author: "Marie Auger-Methe"
date: "28/02/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Document info

This tutorial is heavily based on Appendix S1 (Supporting Information) from Auger-Méthé, M., Newman, K., Cole, D., Empacher, F., Gryba, R., King, A. A., Leos-Barajas, V., Mills Flemming, J., Nielsen, A., Petris, G., and Thomas, L.. 2021. A guide to state–space modeling of ecological time series. Ecological Monographs 91(4):e01470. 10.1002/ecm.1470 (Open access,  https://doi.org/10.1002/ecm.1470). Please refer to the paper and its appendices for more in-depth information on state-space models and how to apply them to data using R.

# Important installation info (to do before seminar!)

Because `TMB` use C++ code, the installation of the package requires additional steps. The package is available on CRAN https://cran.r-project.org/web/packages/TMB/index.html, however your computer needs to be set up so you can compile the C++ code. You can find installation information at: https://github.com/kaskr/adcomp/wiki/Download. Briefly, if you are Mac user, you will need to install the Xcode developer tools (freely available via the App store). With older versions of Mac OS, you may run into problems with the Fortran compiler. For solutions, see the link above and this link https://mac.r-project.org/tools/. If you are a Windows user, you will need to install Rtools: https://cran.r-project.org/bin/windows/Rtools/. You can find guidelines on how to install `TMB` on a Windows machine here: https://github.com/kaskr/adcomp/wiki/Windows-installation.


# Introducing the method with a simple univariate linear Gaussian state-space models

## Model

We start by exploring a very simple SSM: a simple linear SSM with Normal distributions. This model is not linked to an ecological example; it's just a teaching tool. This linear Gaussian SSM
consists of two equations for two time series.

The process equation (or state equation) is:
\begin{equation}
  z_t = \beta z_{t-1} + \epsilon_t, \;\;\; \epsilon_t \sim \text{N}(0, \sigma_p^2),
  \label{E.toy4p.p}
\end{equation}
where $z_t$ is the state value at time $t$, for $t=1, ..., T$ and $\beta$ represents the autocorrelation in the state values. The states are generally unknown, i.e., cannot be observed directly, and are sometimes referred to as latent states or hidden states. This equation represents the evolution of the hidden state as a correlated random walk. The equation implies that there is some stochasticity in the process. This stochasticity is often referred as process variation and is here described by a Normal (Gaussian) distribution with standard deviation $\sigma_p$. For simplicity, we set the initial state value to be 0, i.e., $z_0 =0$.

The observation equation (or measurement equation) is:
\begin{equation}
  y_t = \alpha z_{t} + \eta_t, \;\;\; \eta_t \sim \text{N}(0, \sigma_o^2),
  \label{E.toy4p.o}
\end{equation}
where $y_t$ is the observation at time $t$, for $t=1, ..., T$ and $\alpha$ is a constant of proportionality that represents any systematic discrepancy between the observations and the states (e.g., the average detection rate). This equation links the observation at time $t$ to the underlying state at that time. The equation implies that we are observing the process with error. This error term is often referred as the observation error and, here, is described by a Normal distribution with standard deviation $\sigma_o$.

This model has four parameters: $\alpha, \beta, \sigma_o, \sigma_p$. It is sometimes difficult to accurately estimate all four parameters, something we will demonstrate in the sections below. It is useful to explore a simpler model, where we fix two of the parameters and only have two parameters to estimate. In the simpler model, we fix both $\alpha$ and $\beta$ to 1, resulting in the following.

The process equation is now:
\begin{equation}
  z_t = z_{t-1} + \epsilon_t, \;\;\; \epsilon_t \sim \text{N}(0, \sigma_p^2),
  \label{E.toy2p.p} 
\end{equation}
and represents the evolution of the hidden state as a simple random walk. The observation equation is now:
\begin{equation}
  y_t = z_{t} + \eta_t, \;\;\; \eta_t \sim \text{N}(0, \sigma_o^2),
  \label{E.toy2p.o}
\end{equation}
where we assume that there is no systematic bias in the observation (i.e. we observed the state with some error but the observations are not systematically smaller or larger than the states).

## Simulations

Here, we simulate data using the linear Gaussian SSM described in Eqs. \ref{E.toy2p.p}-\ref{E.toy2p.o} from section \ref{LGSSMm} above. This is the simpler model where two of the parameters are fixed ($\alpha=\beta=1$).

First, let us simulate the process for a time series of length 200 ($T=200$), with an additional time step for the state at $t=0$. To be consistent with the model description, we set to $z_0 = 0$. We choose the standard deviation of the process variation, $\sigma_p$, to be 0.1.

```{r process.sim, tidy=FALSE}
# Create a vector that will keep track of the states
# It's of length T + 1 (+1 for t=0)
# T is not a good name in R, because of T/F, so we use TT
TT <- 200
z <- numeric(TT + 1)
# Standard deviation of the process variation
sdp <- 0.1
# Set the seed, so we can reproduce the results
set.seed(553)
# For-loop that simulates the state through time, using i instead of t,
for(i in 1:TT){
  # This is the process equation
  z[i+1] <- z[i] + rnorm(1, 0, sdp)
  # Note that this index is shifted compared to equation in text,
  # because we assume the first value to be at time 0
}
```

Let us plot the time series we have created.

```{r process.sim.fig}
plot(0:TT, z,
     pch = 19, cex = 0.7, col="red", ty = "o", 
     xlab = "t", ylab = expression(z[t]), las=1)
```

Second, let us simulate the observations. We set the standard deviation of the observation error $\sigma_o$ to 0.1.

```{r obs.sim, tidy=FALSE}
# Create a vector that will keep track of the observations
# It's of length T
y <- numeric(TT)
# Standard deviation of the observation error
sdo <- 0.1
# For t=1, ... T, add measurement error
# Remember that z[1] is t=0
y <- z[2:(TT+1)] + rnorm(TT, 0, sdo)
```

Let us plot both the observations and the states. From now on, we are adding extra space on the y-axis to leave space for the legend. Note that the space we assigned may not work for all figure sizes.

```{r obs.sim.fig, tidy=FALSE}
plot(1:TT, y,
     pch=3, cex = 0.8, col="blue", ty="o", lty = 3,
     xlab = "t", ylab = expression(z[t]),
     xlim = c(0,TT), ylim = c(min(y), max(y)+max(y)/5),
     las = 1)
points(0:TT, z,
       pch = 19, cex = 0.7, col="red", ty = "o")
legend("top",
       legend = c("Obs.", "True states"),
       pch = c(3, 19),
       col = c("blue", "red"),
       lty = c(3, 1),
       horiz=TRUE, bty="n", cex=0.9)
```

With real data, we usually only have the observations, $y_t$, and do not know the true states, $z_t$. However, simulated data allow us to see the discrepancies between the observed values and the values for the process they represent.

## Fitting the model using TMB

In TMB you need to create a C++ file that computes the value of the negative log likelihood for your data and for a given set of parameter values. Then, you compile the C++ code using `TMB` and use that function in one of R's optimizers to find the minimum negative log likelihood. The beauty of `TMB` is that it uses the Laplace approximation to integrate over the states and computes the gradient efficiently, which speeds up the optimizing process.


Assuming you have properly installed \texttt{TMB}, let's load it.

```{R LoadTMB, message=FALSE}
library(TMB)
```

### Writing the negative log-likelihood function in C++

Now, let us create the negative log-likelihood function that we will minimize (equivalent of maximizing the likelihood). We write this function in a special TMB C++ language. The code for the function needs to be save in a text file. We usually give it the extension .cpp. You can do this directly in R Studio. You write the code just as you would write a separate R script, but you save it as a C++ file by giving it the .cpp extension.

We name the file **toy2p.cpp** and it will contain the following code. Note that in C++, comments are preceded by // or contained within /* */. Also, as explained below, here the code is a little bit more complex than strictly necessary. We have included code that allow us to compute the one-step-ahead residuals and simulate from the model. While these features are not always necessary, they are key to model checking and thus we believe they should be part of all model fitting workflow.

```{Rcpp toy2p.cpp, eval=FALSE}
/*----------------------- SECTION A --------------------------*/
// Link to the TMB package
#include <TMB.hpp>

/*----------------------- SECTION B --------------------------*/
// Define main function
template<class Type>
Type objective_function<Type>::operator() ()
{

  /*----------------------- SECTION C --------------------------*/
  // Specify the input data
  DATA_VECTOR(y);

  // For one-step-ahead residuals
  DATA_VECTOR_INDICATOR(keep, y);

  // Specify the parameters
  PARAMETER(logSdP); // Log of st. dev. for the process variation
  PARAMETER(logSdO); // Log of st. dev. for the observation error

  // Specify the random effect/states
  PARAMETER_VECTOR(z);

  /*----------------------- SECTION D --------------------------*/
  // Transform standard deviations
  // exp(par) is a trick to make sure that the estimated sd > 0
  Type sdp = exp(logSdP);
  Type sdo = exp(logSdO);

  /*----------------------- SECTION E --------------------------*/
  // Define the variable that will keep track of
  // the negative log-likelihood (nll)
  Type nll = 0.0;

  /*----------------------- SECTION F --------------------------*/
  // Calculate the contribution to the negative log-likelihood
  // of the process equation for t=1,...,T
  // Remember that we fixed z_0 = 0
  for(int i = 1; i < z.size(); ++i){
    nll -= dnorm(z(i), z(i-1), sdp, true);

    //*----------------------- SECTION G --------------------------*/
    // Simulation block for process equation
    SIMULATE {
      z(i) = rnorm(z(i-1), sdp);
    }
  }

  /*----------------------- SECTION H --------------------------*/
  // Calculate the contribution to the negative log-likelihood
  // of the observation equation for t=1,...,T
  // Remember, the first element of z is at t=0,
  // while the first element of y is at t=1
  for(int i = 0; i < y.size(); ++i){
    nll -= keep(i)*dnorm(y(i), z(i+1), sdo, true);

    //*----------------------- SECTION I --------------------------*/
    // Simulation block for observation equation
    SIMULATE {
      y(i) = rnorm(z(i+1), sdo);
    }
  }


  /*----------------------- SECTION J --------------------------*/
  // State the transformed parameters to report
  // Using ADREPORT will return the point values and the standard errors
  // Note that we only need to specify this for parameters
  // we transformed, see section D above
  // The other parameters, including the random effects (states),
  // will be returned automatically
  ADREPORT(sdp);
  ADREPORT(sdo);

  /*----------------------- SECTION K --------------------------*/
  // Report simulated values
  SIMULATE{
    REPORT(z);
    REPORT(y);
  }


  /*----------------------- SECTION L --------------------------*/
  // State that we want the negative log-likelihood to be returned
  // This is what the optimizer in R will minimize
  return nll;

}

```


A few notes about C++:

- C++ uses round brackets `()` for both objects and functions

- the index of an object in C++ starts at 0, not 1. For example, to get the first element of a vector called `x`, you write `x(0)`. This is why the for-loops in sections F and H end before the length of the index, which is specified by `.size()`. If R crashes when you compile the code (see below for compiling instruction), the most likely culprit is an index problem, so check your for-loops.

- lines end with `;`

- Many of the density functions available in R are available in C++, at least when using the `TMB package`. These usually have the same names and arguments as those in R. For example, in sections F and H we use `dnorm` just like we would in R.

A few tips about writing your function:

- If we want to use an unconstrained optimizer in R, we have to specify in the negative log-likelihood function that we transform the values provided by the optimizer into values that are in the appropriate domain. For example, here in  section D, we exponentiate the input log-transformed standard deviations (logSdP and logSdO) to ensure that the standard deviations have positive values. We do this so that we can use an unconstrained optimizer, yet only have positive estimated values for the standard deviations.

- By default, `TMB` will only return the estimated values in the same form as given in the function (e.g., `logSdP` rather than `sdp`). To be able to retrieve the estimate values of interest, we need to specify that we want them returned. This is done in section J.

- Note that we have included simulation blocks (sections G, I, and K), even though these blocks are not necessary to fit the model to data. These blocks are extremely useful for model validation, see the model validation section below for details.

- Similarly, while it is not necessary to fit the model, we have added an indicator variable (`keep`) in both sections C and H. This indicator variable is necessary to calculate the one-step-ahead residuals that we will discuss in the model validation section.

Now that a few things are defined, it may be easier to go back to the function and see what it means. 

- Section A loads the packages needed. 

- Section B indicates that we are defining the main function that will calculate the negative log likelihood. 

- Section C defines the data and parameters. The values of both data and parameters will be put in through R. 

- Section D transforms the parameters so that they are constrained appropriately (here, that the standard deviation are $>0$). 

- Section E creates the object that will keep track of the negative log likelihood. 

- Section F has a for-loop that calculates the contribution to the negative log likelihood of the process equation. 

- Section G allows to simulate the process equation.

- Section H calculates the contribution to the negative log likelihood of the observation equation. Note that using \texttt{keep} tells \texttt{TMB} that this is where we would remove the data for the one-step ahead residuals. 

- Section I allows to simulate the observation equation. 

- Section J indicates that we want the parameters in their untransformed form to be returned with their standard errors. 

- Section K indicates what to report when we simulate the model. 

- Section L indicates that the main item the function returns is the negative log likelihood.

Now that we have the negative log likelihood C++ file saved, we go back to R.

### Compiling and loading C++ function

The first thing we want to do is compile the C++ code and load it so we can use it to calculate the negative log likelihood of our model. Here, our **toy2p.cpp** file is found in the working directory.

```{r tmbCompile,  results="hide"}
compile("toy2p.cpp", flags="-Wno-unused-variable")
dyn.load(dynlib("toy2p"))
```

### Preparing the model for fitting

First, we need to prepare the data. This is going to be a list with all the items found in the .cpp files that are of the type `DATA` (e.g., `DATA\_VECTOR`). Here, we only have \texttt{y} which is the time series of observed values.

```{r prepY}
dataTmb <- list(y = y)
```

Second, we need a list with the parameters. Again the names need to match those in the .cpp file. These are the starting values for the parameters. Note that this includes both the parameters and the states.

```{r parPrep, tidy=FALSE}
par2pTmb <- list(logSdP = 0, logSdO = 0,
             z=rep(0,length(dataTmb$y)+1))
```

By default all values of `par2pTmb` will be estimated, but we assume that we know the value of the state at time 0, $z_0 = 0$. To provide this information, we can create a named list that we will input for the argument `map` of the function `MakeADFun` below. The elements of `map` should have the same name and size as those in `par2pTmb`, but we only need to specify the elements that we want to manipulate, here `par2pTmb\$z`. To fix a value, we set that value to `NA`. The values that are estimated should have a different entry and all values should be factors.

```{r tmbMapPrep}
mapTmb <- list(z=as.factor(c(NA,1:length(dataTmb$y))))
```

Before we can fit the model, we need to use the function `MakeADFun` to combine the data, the parameters, and the model and create a function that calculates the negative
log likelihood and the gradients. 

To identify our random effects, namely the states, $z_t$, we set the argument `random` to equal the name of the parameters that are random effects. The argument `DLL` identify the compiled C++ function to be linked.

```{r MakeADFun, tidy=FALSE}
m2pTmb <- MakeADFun(data=dataTmb, 
                    parameters=par2pTmb, 
                    map=mapTmb, 
                    random= "z",
                    DLL= "toy2p")
```

### Fitting the model to data

We fit the model using \texttt{nlminb}, which is a base R optimizer, but we input the object returned by `MakeADFun`.

```{r tmbOpt}
f2pTmb <- nlminb(start = m2pTmb$par, # Initial values for the parameters
                 objective = m2pTmb$fn, # Function to be minimized
                 gradient = m2pTmb$gr) # Gradient of the objective
```

It is important to check whether the model converged.

```{r}
f2pTmb$message
```

A message stating \texttt{"both X-convergence and relative convergence (5)"} would also indicate convergence.

### Exploring the results

To look at the parameter estimates, you can use the function `sdreport` and object object created by `MakeADFun`.

```{r tmbEst, tidy=FALSE}
sdr2pTmb <- summary(sdreport(m2pTmb))
# This will have the parameters and states
# So we can just print the parameters
sdr2pTmb[c("sdp", "sdo"),]
```

We can see that the estimates for both the process variation and the observation error are close to their true value of 0.1, with relatively small standard errors (SEs).

We can get the smoothed state values with the `sdreport` function as above and this will give you the SEs for these states. If you are only interested in the state values (not theis SEs), you can also extract them directly from the model object.

```{r}
# To get the point estimate and the SE of the states
zsSe2pTmb <- sdr2pTmb[row.names(sdr2pTmb)=="z",]
head(zsSe2pTmb)
# To get only the point estimates of the states
zs2pTmb <- m2pTmb$env$parList()$z
head(zs2pTmb)
```

Note that in this case, because we fixed the value of the state at time $t=0$, the first method, which looks at the predicted values, only returns state values for $t>0$.

We can use the SEs to calculate the 95\% confidence intervals.

```{r tmbCI, tidy=FALSE}
zsCIl2pTmb <- zsSe2pTmb[,1] + 
  qnorm(0.025, sd = zsSe2pTmb[,2])
zsCIu2pTmb <- zsSe2pTmb[,1] + 
  qnorm(0.975, sd = zsSe2pTmb[,2])
```

We can now overlay the state estimates along with their confidence intervals on the plot of the simulated data.

```{R tmbRp, tidy=FALSE}
plot(1:TT, y,
     pch=3, cex=0.8, col="blue", ty="o", lty = 3,
     xlab = "t", ylab = expression(z[t]),
     xlim = c(0,TT), ylim = c(min(y), max(y)+max(y)/5),
     las = 1)
points(0:TT, z,
       pch=19, cex=0.7, col = "red", ty="o")
polygon(c(1:TT, TT:1), c(zsCIl2pTmb,rev(zsCIu2pTmb)),
        col=rgb(1,0.7,0.4,0.3), border=FALSE)
lines(0:TT, zs2pTmb, 
      col= "darkgoldenrod1", lwd = 2)
legend("top",
       legend = c("Obs.", "True states", "Smooth. states"),
       pch = c(3, 19, NA),
       col = c("blue", "red", "darkgoldenrod1"),
       lwd = c(1, 1, 2), lty = c(3, 1, 1),
       horiz=TRUE, bty="n", cex=0.9)
```

### Model selection

We sometimes have more than one model to explore. For example, we could fit the more general and slightly more complex version of the model, where we are estimating in addition to the standard
deviations, two additional parameter: $\alpha$ and $\beta$. 

The process equation of this new model is:
\begin{equation}
  z_t = \beta z_{t-1} + \epsilon_t, \;\;\; \epsilon_t \sim \text{N}(0, \sigma_p^2),
  \label{E.toy4p.p}
\end{equation}
where $\beta$ represents the autocorrelation in the state values. 

The observation equation of this new model is:
\begin{equation}
  y_t = \alpha z_{t} + \eta_t, \;\;\; \eta_t \sim \text{N}(0, \sigma_o^2),
  \label{E.toy4p.o}
\end{equation}
where $\alpha$ is a constant of proportionality that represents any systematic discrepancy between the observations and the states (e.g., the average detection rate).

In our simulation, both $\alpha$ and $\beta$ are implicitly set to 1. Here we estimate these values.

To fit this model in ```TMB```, we will create a new file called **toy4p.cpp**. Note that this is exactly the same as **toy2p.cpp** except that we added the two new parameters in sections C, and F to I.

```{Rcpp toy4p.cpp, write_chunk=TRUE, eval=FALSE}
/*----------------------- SECTION A --------------------------*/
// State that we need the TMB package
#include <TMB.hpp>

/*----------------------- SECTION B --------------------------*/
// Define main function
template<class Type>
Type objective_function<Type>::operator() ()
{

  /*----------------------- SECTION C --------------------------*/
  // Specify the input data
  DATA_VECTOR(y);

  // For one-step-ahead residuals
  DATA_VECTOR_INDICATOR(keep, y);

  // Specify the parameters
  PARAMETER(logSdP); // Log of st. dev. for the process variation
  PARAMETER(logSdO); // Log of st. dev. for the observation error
  PARAMETER(alpha);
  PARAMETER(beta);

  // Specify the random effect/states
  PARAMETER_VECTOR(z);

  /*----------------------- SECTION D --------------------------*/
  // Transform standard deviations
  // exp(par) is a trick to make sure that the estimated sd > 0
  Type sdp = exp(logSdP);
  Type sdo = exp(logSdO);

  /*----------------------- SECTION E --------------------------*/
  // Define the variable that will keep track of
  // the negative log-likelihood (nll)
  Type nll = 0.0;

  /*----------------------- SECTION F --------------------------*/
  // Calculate the contribution to the negative log-likelihood
  // of the process equation for t=1,...,T
  // Remember that we fixed z_0 = 0
  for(int i = 1; i < z.size(); ++i){
    nll -= dnorm(z(i), beta*z(i-1), sdp, true);

    //*----------------------- SECTION G --------------------------*/
    // Simulation block for process equation
    SIMULATE {
      z(i) = rnorm(beta*z(i-1), sdp);
    }
  }

  /*----------------------- SECTION H --------------------------*/
  // Calculate the contribution to the negative log-likelihood
  // of the observation equation for t=1,...,T
  // Remember, the first element of z is at t=0,
  // while the first element of y is at t=1
  for(int i = 0; i < y.size(); ++i){
    nll -= keep(i)*dnorm(y(i), alpha*z(i+1), sdo, true);

    //*----------------------- SECTION I --------------------------*/
    // Simulation block for observation equation
    SIMULATE {
      y(i) = rnorm(alpha*z(i+1), sdo);
    }
  }

  /*----------------------- SECTION J --------------------------*/
  // State the transformed parameters to report
  // Using ADREPORT will return the point values and the standard errors
  // Note that we only need to specify this for parameters
  // we transformed, see section D above
  // The other parameters, including the random effects (states),
  // will be returned automatically
  ADREPORT(sdp);
  ADREPORT(sdo);

  /*----------------------- SECTION K --------------------------*/
  // Report simulated values
  SIMULATE{
    REPORT(z);
    REPORT(y);
  }


  /*----------------------- SECTION L --------------------------*/
  // State that we want the negative log-likelihood to be returned
  // This is what the optimizer in R will minimize
  return nll;

}

```

As before, we compile and load the function.

```{r tmbfCompile, results="hide"}
compile("toy4p.cpp", flags="-Wno-unused-variable")
dyn.load(dynlib("toy4p"))
```

We can use the same data and map objects as above, but we need to change the object containing the initial parameter values, because we have two new parameters to estimate. 

We then get the MLE. Note that while we can often use uninformative starting values (e.g., 0), we can run into estimation problems if we set all the starting values as 0s. To avoid these problems, we use the true simulated values for the starting values. With real data, we could have started with sensible values rather than 0.

```{r tmbfMLE, tidy=FALSE}
par4pTmb <- list(logSdP = log(0.1), logSdO = log(0.1), 
                 alpha = 1, beta = 1, # New parameters
                 z=rep(0,length(dataTmb$y)+1))
m4pTmb <- MakeADFun(dataTmb, par4pTmb, map=mapTmb, random= "z",
                  DLL= "toy4p", silent=TRUE)
f4pTmb <- nlminb(m4pTmb$par, m4pTmb$fn, m4pTmb$gr)
f4pTmb$message
```

We can see here that the optimizer has converged.

We can use again `sdreport` to get the parameter estimates and their SEs.

```{r tmbfest, tidy=FALSE}
sdr4pTmb <- summary(sdreport(m4pTmb))
cbind(Simulated = c(sdp, sdo, 1, 1),
      sdr4pTmb[c("sdp", "sdo", "beta", "alpha"),])
```

We can see here that compared to the estimates from the original model, the more complex and flexible model has much
higher standard errors. As mentioned above, the optimization process is also quite sensitive to the starting values.

We can use AIC to compare these two models. Since the \texttt{C++} function calculates the negative log likelihood and is the objective of the optimization procedures, we can just use the returned value by `nlminb`.

```{r tmbAIC, tidy=FALSE}
aic2pTmb <- 2 * 2 + 2 * f2pTmb$objective
aic4pTmb <- 2 * 4 + 2 * f4pTmb$objective
c(Original = aic2pTmb, 
  Flexible = aic4pTmb,
  Difference = aic4pTmb - aic2pTmb)
```

We can see that the original model has a lower AIC than the more flexible model. Thus the original model is considered better. If the two models were equivalent, we would expect a AIC difference of 4 ($2k$), but here the difference is much smaller: \Sexpr{round(aic4pTmb - aic2pTmb,2)}. This may be because the more flexible model is overfitting the data.
